---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# DIMPLE: **D**istance based **I**nference for **M**ultip**L**ex imaging **E**xperiments


<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/DIMPLE)](https://CRAN.R-project.org/package=DIMPLE)
[![codecov](https://codecov.io/gh/nateosher/DIMPLE/branch/master/graph/badge.svg?token=87UMWJ42XU)](https://codecov.io/gh/nateosher/DIMPLE)
<!-- badges: end -->

The goal of DIMPLE is to provide tools and infrastructure to facilitate the manipulation, exploration, and modeling of multiplex imaging data.

## Installation

DIMPLE is available from GitHub. We recommend using `pak` so the package and its required R package dependencies are installed together:

``` r
# Install pak from CRAN if it is not already installed
install.packages("pak")

# Install DIMPLE and its dependencies
pak::pkg_install("bayesrx/DIMPLE")
```

### Run the Shiny app locally

After installation, open R or RStudio and run:

``` r
library(DIMPLE)
shiny_app(full_app = TRUE)
```

The application opens in a web browser and runs locally on your computer; keep the R session open while using it. The full local app enables raw-data quantification and spatial-distance computation that are disabled in the lightweight hosted version. An internet connection is required for installation and the first use of the lung cancer example dataset. After DIMPLE and the example data are installed or cached, the app can run locally without repeatedly downloading those resources.

The Cohort-level summary tab also includes the restored pairwise regression screen below the descriptive cohort plot, with optional covariate and cell-count adjustment and an FDR-adjusted regression heatmap.

## Example

Suppose we're interested in exploring the lung cancer data from the [`VectraPolarisData`](https://bioconductor.org/packages/release/data/experiment/html/VectraPolarisData.html) package. Assuming you've already installed that package, we can build a `MltplxExperiment` object, the core of `DIMPLE`, from the data as follows:

```{r}
library(DIMPLE)
library(tidyverse)
lung_data = VectraPolarisData::HumanLungCancerV3()

# Get the x coordinates for the cells
cell_x_coords = SpatialExperiment::spatialCoords(lung_data)[,1] %>% as.numeric()
# Get the y coordinates for the cells
cell_y_coords = SpatialExperiment::spatialCoords(lung_data)[,2] %>% as.numeric()
# Get the patient ids for the slides
patient_ids = lung_data$slide_id
# Get the slide ids for the slides
slide_ids = lung_data$sample_id

# Map phenotypes to character vector - there are more efficient ways to do this,
# but this way is the most legible
test_n = length(lung_data$phenotype_cd14)
tictoc::tic()
cell_marks = tibble(
  cd_14 = lung_data$phenotype_cd14[1:test_n] == "CD14+",
  cd_19 = lung_data$phenotype_cd19[1:test_n] == "CD19+",
  cd_4 = lung_data$phenotype_cd4[1:test_n] == "CD4+",
  cd_8 = lung_data$phenotype_cd8[1:test_n] == "CD8+",
  ck = lung_data$phenotype_ck[1:test_n] == "CK+",
  other = lung_data$phenotype_other[1:test_n] == "Other+"
) %>%
  as.matrix() %>%
  (\(m){
    m %*% 1:6
  }) %>%
  map_chr(\(x) c(NA, "CD14", "CD19", "CD4", "CD8", "CK", "Other")[x + 1])

na_marks = which(is.na(cell_marks))

na_marks = which(is.na(cell_marks))
cell_x_coords = cell_x_coords[-na_marks]
cell_y_coords = cell_y_coords[-na_marks]
patient_ids = patient_ids[-na_marks]
slide_ids = slide_ids[-na_marks]
cell_marks = cell_marks[-na_marks]

lung_experiment = DIMPLE::new_MltplxExperiment(
  x = cell_x_coords,
  y = cell_y_coords,
  marks = cell_marks,
  slide_id = slide_ids
)

lung_experiment
```

This object can be indexed like a list- each entry is of class `MltplxObject`, and stores data about a particular biopsy:

```{r }
lung_experiment[[1]]
```

Using the implemented generic functions, this allows for quick inspection of a given biopsy:

```{r example_plot}
plot(lung_experiment[[1]])
```

To learn more about these data structures and how to utilize them for exploration, check out the `mltplx-experiment` vignette. `DIMPLE` also provides functions for simulating Multiplex Imaging data- an overview of this functionality can be found in the `simulations` vignette

